---
title: "PS270 Final Project"
author: "Katie Troske"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to statistics about literacy and education across different areas and community types, and how they compare to voting and elections. I think it would be interesting to study the correlation between literacy levels and political party affiliation/candidate preference.

## Research Question and Hypothesis

Question to be researched and studied in the data: Did higher education levels in a county lead to more support for Democratic candidates in the 2020 Presidential Elections?

My hypothesis: I hypothesize that counties with more adults holding a bachelor’s degree or higher will vote more for Democrats, because education is typically linked to political engagement and more liberal social views. 

## Description of Data Set and Research Question (paragraph for milestone 2)

For my project, I will use county level data from the U.S. Census and 2020 presidential election results. My research question is: **“Did higher education levels in a county lead to more support for Democratic candidates in the 2020 Presidential Elections?”**. I hypothesize that counties with more adults holding a bachelor’s degree or higher will vote more for Democrats, because education is linked to political engagement and more liberal social views. My main explanatory variable will be education level, measured as the percentage of adults in a county with a bachelor’s degree or higher. This will vary across counties and states, allowing us to observe how variation impacts the dependent/outcome variable. The outcome variable will be Democratic vote share, which is measured as the percentage of total votes cast for the Democratic candidate in each county. If the patterns in the data show a positive relationship between higher education and democratic vote, my hypothesis is supported. However, if the relationship is zero or negative, rather showing a correlation between Republican vote and education, it is not supported.


## Datasets being used for project and code to load it

For my project, I am using a GitHub dataset that I have loaded under "election_data" that contains election results by county from the 2020 election cycle. To understand education and literacy levels, I will use the "tidycensus" package that has census and education data by county. 

```{r setup, include = FALSE}
library(tidycensus)
library(readr)
library(dplyr)
library(tidyr)
```


```{r Loading the data}

election_data <- read_csv("https://raw.githubusercontent.com/tonmcg/US_County_Level_Election_Results_08-20/master/2020_US_County_Level_Presidential_Results.csv")

head(election_data)
```

## Cleaning and Visualizing my data (Milestone 3)

```{r stepup chunk}
library(tidycensus)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

```

Accessing education levels from the census using an API key 

```{r}
census_api_key("2de76ac2174e4c5d914da47a19564a5de7799580", install = TRUE, overwrite = TRUE)
```

Organize the values and information we want to pull from the census dataset

```{r}
edu_data <- get_acs(
  geography = "county",
  variables = c(
    bachelor = "B15003_022",
    master = "B15003_023",
    professional = "B15003_024",
    doctorate = "B15003_025",
    total = "B15003_001"
  ),
  year = 2020,
  survey = "acs5",
  geometry = FALSE
)
```

Clean the education data to be more focused and interpretable, selecting GEOID (county) and percent of the population with a bachelor's degree or higher 

```{r}
edu_clean <- edu_data |>
  select(GEOID, variable, estimate) |>
  pivot_wider(names_from = variable, values_from = estimate) |>
  mutate(
    bachelors_pct = (bachelor + master + professional + doctorate) / total * 100
  ) |>
  select(GEOID, bachelors_pct)

head(edu_clean)
```

Clean election data by focusing on the percentage of democratic vote share and GEOID

```{r}
election_clean <- election_data |>
  mutate(
    total_votes = votes_dem + votes_gop,
    dem_share = votes_dem / total_votes,
    GEOID = county_fips
  ) |>
  select(GEOID, dem_share)

```


Merge the datasets together using left_join and connecting the sets via the GEOID character 


```{r}
merged_data <- left_join(election_clean, edu_clean, by = "GEOID")

head(merged_data)
```

Produce a scatter plot that represents the relationship between democratic vote share and education rates by county 

```{r}
ggplot(merged_data, aes(x = bachelors_pct, y = dem_share)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "pink") +
  labs(
    title = "County Education vs Democratic Vote Share (2020)",
    x = "Percent with Bachelor's Degree or Higher",
    y = "Democratic Vote Share"
  ) 
```

## Regression Analysis
**The relationship between the percent of those with Bachelor's degrees and Democratic vote share** 

The scatterplot and regression line show a **positive relationship between educational level by county and Democratic vote share in the 2020 election**. Each point represents a county, with the x-axis showing the percentage of residents who hold a bachelor’s degree or higher and the y-axis showing the Democratic share of the vote. The upward slope of the regression line indicates that counties with higher levels of education tended to give a larger share of votes to the Democratic candidate. In general, counties with lower percentages of college-educated residents were more likely to have lower Democratic vote shares, while counties with more highly educated populations were more likely to vote Democratic. However, the data points are somewhat spread out around the line, showing that education alone does not fully explain voting patterns. Other factors such as income, race, region, urbanization, and local political culture likely also influenced election outcomes. **Overall, the regression suggests that educational attainment was an important predictor of Democratic vote share in 2020**.


## Analyzing Visuals of my Data (Milestone 4)

Producing a Boxplot with education groups to examine and visualize relationship between education and voting. 

Step One: Creating Education Groups

```{r}
merged_data <- merged_data |>
  mutate(
    edu_group = ntile(bachelors_pct, 3))

merged_data <- merged_data |>
  mutate(
    edu_group = factor(edu_group,
                       levels = c(1, 2, 3),
                       labels = c("Low Education", "Medium Education", "High Education"))
  )
```

Step Two: Create Boxplot

```{r}
ggplot(merged_data, aes(x = edu_group, y = dem_share)) +
  geom_boxplot(fill = "pink", color = "deeppink") +
  labs(
    title = "Democratic Vote Share by Education Level Groups",
    x = "Education Level",
    y = "Democratic Vote Share"
  ) +
  theme_minimal()
```

## Table: Democratic Vote Share in Relation to Education Level

Creating a table that helps further analyze the relationship between education level and Democratic vote share

```{r}
merged_data |>
  group_by(edu_group) |>
  summarize(avg_dem_share = mean(dem_share, na.rm = TRUE),
    n = n()) |>
  rename(
    "Education Level" = edu_group,
    "Avg. Democratic Vote Share" = avg_dem_share,
    "Number of Counties" = n) |>
  knitr::kable(digits = 3, caption = "Average Democratic Vote Share by Education Group")
```

## Analysis and Discussion

The results of both the boxplot and the table above show a **positive relationship between higher education level and Democratic vote share**. The table illustrates that counties with higher levels of education have higher average Democratic vote shares, with 0.281 in low education counties but 0.438 in high education counties. This pattern is also observed in the boxplot, which depicts higher Democratic support as education increases. 

**Overall, the data supports my hypothesis that more educated counties tended to vote more Democratic in the 2020 election**.